Comparing R packages for calculation of persistent homology

### 2020-02-20 tdabench-paper/tdabench.Rmd
resolved remaining merge conflict and shortened initial paragraph

\section{Methods}
<!--An exclusively R (v3.6.2) code pipeline was built for this project \citep{Rlang,ihaka1996}.-->
We use \CRANpkg{readr} v1.3.1 to read rectangular data \citep{readr}, \CRANpkg{ggplot2} v3.2.1 \citep{ggplot2}, \CRANpkg{scatterplot3d} v0.3-41 \citep{scatterplot3d}, \CRANpkg{recexcavAAR} v0.3.0 \citep{recexcavAAR}, and \CRANpkg{magick} v2.2 \citep{magick} to visualize data, \CRANpkg{bench} v1.0.4 to collect benchmark data \citep{bench}, \CRANpkg{TDA} v1.6.9 \citep{TDA} and \CRANpkg{TDAstats} v0.4.1 \citep{TDAstats} to calculate persistent homology of Vietoris-Rips and alpha simplicial complices, and \CRANpkg{pryr} v0.1.4 for calculations involving R objects \citep{pryr}.
Runtime calculations were averaged over 10 iterations and are visualized as mean $\pm$ SD.
Datasets were generated by sampling functions in base R to generate points uniformly distributed over circles (dimension = 2), spheres (dimension = 3), filled squares (dimension = 2), filled cubes (dimension = 3, 4), and tori (dimension = 3).
The number of points per point cloud varied from 25 to 500 along intervals of 25 points, which were empiric limits chosen after considering available computational resources.
For consistency between software libraries, the minimum and maximum simplicial complex radii were predetermined for each point cloud and provided as parameters to the \CRANpkg{TDA} and \CRANpkg{TDAstats} packages.
Within the \CRANpkg{TDA} package, benchmark data was collected for the GUDHI \citep{gudhi} and Dionysus \citep{Dionysus} libraries; within the \CRANpkg{TDAstats} package, benchmark data was collected for the Ripser \citep{Ripser} library.
As alpha complex calculation was only implemented in GUDHI, alpha complex benchmark data was naturally only collected for the single library.
Measuring memory usage proved challenging since all the libraries calculating persistent homology were implemented in either C++ or Java and then wrapped in R as part of a CRAN package.
Thus, memory burden was indirectly measured by using boundary matrix size as a proxy.
Given that Ripser optimizes computation of persistent homology by avoiding calculation of a boundary matrix, memory use benchmarks are not provided for Ripser and, consequently, \CRANpkg{TDAstats}.

@@

Computing persistent homology of a canonical torus grants quick insight into efficiency of each library (Figure \ref{fig:tor}).
Dionysus exhibits the longest average runtime, and, although Ripser and GUDHI have similar runtimes for smaller point clouds, as the number of points increases Ripser eventually has a significant lead.

>>>>

\section{Methods}
<!--An exclusively R (v3.6.2) code pipeline was built for this project \citep{Rlang,ihaka1996}.-->
We use \CRANpkg{readr} v1.3.1 to read rectangular data \citep{readr}, \CRANpkg{ggplot2} v3.2.1 \citep{ggplot2}, \CRANpkg{scatterplot3d} v0.3-41 \citep{scatterplot3d}, \CRANpkg{recexcavAAR} v0.3.0 \citep{recexcavAAR}, and \CRANpkg{magick} v2.2 \citep{magick} to visualize data, \CRANpkg{bench} v1.0.4 to collect benchmark data \citep{bench}, \CRANpkg{TDA} v1.6.9 \citep{TDA} and \CRANpkg{TDAstats} v0.4.1 \citep{TDAstats} to calculate persistent homology of Vietoris-Rips and alpha simplicial complices, and \CRANpkg{pryr} v0.1.4 for calculations involving R objects \citep{pryr}.
Runtime calculations were averaged over 10 iterations and are visualized as mean $\pm$ SD.
Datasets were generated by sampling functions in base R to generate points uniformly distributed over circles (dimension = 2), spheres (dimension = 3), filled squares (dimension = 2), filled cubes (dimension = 3, 4), and tori (dimension = 3).
The number of points per point cloud varied from 25 to 500 along intervals of 25 points, which were empiric limits chosen after considering available computational resources.
For consistency between software libraries, the minimum and maximum simplicial complex radii were predetermined for each point cloud and provided as parameters to the \CRANpkg{TDA} and \CRANpkg{TDAstats} R packages.
Within the \CRANpkg{TDA} package, benchmark data was collected for the GUDHI \citep{gudhi} and Dionysus \citep{Dionysus} libraries; within the \CRANpkg{TDAstats} package, benchmark data was collected for the Ripser \citep{Ripser} library.
As alpha complex calculation was only implemented in GUDHI, alpha complex benchmark data was naturally only collected for the single library.
Measuring memory usage proved challenging since all the libraries calculating persistent homology were implemented in either C++ or Java and then wrapped in R as part of a CRAN package.
Thus, memory burden was indirectly measured by using boundary matrix size as a proxy.
Given that Ripser optimizes computation of persistent homology by avoiding calculation of a boundary matrix, memory use benchmarks are not provided for Ripser and, consequently, \CRANpkg{TDAstats}.

@@

Computing persistent homology of a canonical torus grants quick insight into efficiency of each library (Figure \ref{fig:tor}).
Dionysus exhibits the longest average runtime, and, although Ripser and GUDHI have similar runtimes for smaller point clouds, as the number of points increases Ripser eventually has a significant lead.
Next, we compare performance of these three libraries with multiple canonical datasets to ensure that the noted pattern generalizes.
